---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ALEA-R

ALEA-R is an R package for hydrological frequency analysis.

It provides tools for exploratory summaries, probability distribution fitting,
return-level estimation, bootstrap confidence intervals, goodness-of-fit
assessment, sample diagnostics, AI-assisted distribution-selection support,
batch analysis, publication-ready plots, and export helpers.

## Installation

You can install the development version from GitHub with:

```{r installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("hydrostat/ALEA")
```


## Main features

ALEA-R supports:

- exploratory summaries for hydrological samples;
- fitting of probability distributions used in the FADS_AI selection study;
- return-level estimation;
- percentile bootstrap confidence intervals for return levels;
- goodness-of-fit statistics and information criteria;
- sample diagnostics for data-quality and frequency-analysis assumptions;
- AI-assisted distribution-selection support;
- batch analysis for multiple stations or sites;
- ggplot-based plotting methods;
- plot and table export helpers.

## Supported distributions

The initial implementation supports six candidate distributions:

| Code | Distribution |
|---|---|
| `gev` | Generalized Extreme Value |
| `gpa` | Generalized Pareto |
| `pe3` | Pearson type III |
| `ln2` | Two-parameter lognormal |
| `ln3` | Three-parameter lognormal |
| `gum` | Gumbel |

LP3 is not included in the initial implementation.

## Basic workflow

```{r basic-workflow, eval = FALSE}
library(ALEA)

x <- c(
  42.1, 38.5, 51.3, 47.0, 62.4,
  55.2, 49.8, 58.1, 60.3, 45.7
)

fit <- alea_fit(
  x,
  distribution = "gev",
  method = "lmom"
)

fit
coef(fit)
```

## Return levels

```{r return-levels, eval = FALSE}
rl <- alea_return_level(
  fit,
  return_period = c(10, 25, 50, 100)
)

rl
plot(rl)
```

## Bootstrap confidence intervals

```{r bootstrap-ci, eval = FALSE}
ci <- confint(
  fit,
  parm = "return_level",
  return_period = c(10, 25, 50, 100),
  level = 0.95,
  method = "bootstrap",
  n_boot = 500,
  seed = 123
)

ci
plot(ci)
```

The initial implementation provides percentile bootstrap confidence intervals
for return levels. Parameter confidence intervals, asymptotic return-level
intervals, and generic delta-method intervals are not implemented in the
initial release.

## Goodness-of-fit and diagnostics

```{r gof-diagnostics, eval = FALSE}
gof <- alea_gof(fit)
gof
plot(gof, type = "statistic")

diagnostics <- alea_diagnostics(fit)
diagnostics
plot(diagnostics, type = "status")
```

Goodness-of-fit results report empirical distribution function statistics and
information criteria. Calibrated goodness-of-fit p-values and chi-square
goodness-of-fit tests are deferred.

Diagnostics are sample-level checks intended to flag possible data-quality
issues or assumption concerns. Diagnostic warnings do not automatically
invalidate a fitted model.

## AI-assisted distribution selection

```{r ai-selection, eval = FALSE}
selection <- alea_select(x)

selection
as.data.frame(selection)
plot(selection)
```

ALEA-R includes AI-assisted distribution-selection support through the bundled
FADS_AI lightweight operational application model.

FADS_AI output should be interpreted as model-based decision-support evidence
for candidate distribution families. It is not proof of the true generating
distribution and should not replace goodness-of-fit assessment, diagnostics,
return-level uncertainty evaluation, or hydrological judgement.

## Batch analysis

```{r batch-analysis, eval = FALSE}
batch <- alea_batch_fit(
  data = annual_maxima,
  station = "station",
  time = "year",
  value = "value",
  distributions = c("gev", "gpa", "pe3", "ln2", "ln3", "gum"),
  methods = c("lmom"),
  return_period = c(10, 25, 50, 100),
  gof = TRUE,
  diagnostics = TRUE,
  select = "ai"
)

alea_results(batch, "stations")
alea_results(batch, "fits")
alea_results(batch, "selected_models")
alea_results(batch, "return_levels")
alea_results(batch, "gof")
alea_results(batch, "diagnostics")
alea_results(batch, "errors")
```

Batch workflows use structured error capture. A failure for one station,
distribution, method, return-level calculation, goodness-of-fit calculation,
diagnostic calculation, or selection step does not stop the full workflow.

## Plotting and export

All plot methods return `ggplot` objects.

```{r plots, eval = FALSE}
p <- plot(fit, type = "return_level")

alea_save_plot(
  p,
  filename = "return_level_plot.png",
  width = 7,
  height = 5,
  dpi = 300
)
```

ALEA-R can also export data frames and flat batch result tables:

```{r export, eval = FALSE}
alea_export(rl, path = "return_levels.csv")
alea_export(batch, path = "batch_results", type = "all")
```

## Current limitations

The initial implementation does not include:

- LP3;
- Portuguese API aliases;
- calibrated goodness-of-fit p-values;
- chi-square goodness-of-fit tests;
- parameter confidence intervals;
- asymptotic or delta-method return-level confidence intervals;
- HidroWeb data access in the core package.

## Citation

A formal package citation will be added before the first public release.

```{r citation, eval = FALSE}
citation("ALEA")
```