---
title: "Batch analysis with ALEA-R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch analysis with ALEA-R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Batch analysis with ALEA-R

This vignette shows how to use `alea_batch_fit()` to run the same frequency-analysis workflow for multiple stations or sites.

Batch analysis is useful when the same candidate distributions, estimation methods, return periods, goodness-of-fit statistics, diagnostics, and AI-assisted selection workflow must be applied consistently across several hydrological samples.

The examples use a small synthetic data set so that the vignette can be rebuilt quickly during package checks.

```{r setup}
library(ALEA)
```

# Example data

The input data must be a data frame. The station identifier and observed value columns are required. A time column is optional, but recommended when available.

```{r example-data}
annual_maxima <- data.frame(
  station = rep(c("A", "B", "C"), each = 16),
  year = rep(2001:2016, times = 3),
  value = c(
    42.1, 38.5, 51.3, 47.0, 62.4, 55.2, 49.8, 58.1,
    60.3, 45.7, 53.5, 57.6, 64.2, 59.1, 61.5, 66.0,
    75.4, 80.2, 69.5, 91.3, 88.1, 84.6, 92.0, 86.7,
    95.5, 90.1, 83.8, 97.2, 101.3, 94.7, 98.5, 103.4,
    25.2, 29.4, 31.1, 28.7, 34.5, 32.0, 35.2, 30.8,
    33.9, 37.4, 36.1, 39.2, 38.5, 40.0, 41.3, 43.1
  )
)

head(annual_maxima)
```

# Run a batch workflow

The following workflow fits all six initial ALEA-R candidate distributions to each station using L-moment estimation. It also computes quantiles, goodness-of-fit results, sample diagnostics, and AI-assisted distribution-selection support.

```{r batch-fit}
batch <- alea_batch_fit(
  data = annual_maxima,
  station = "station",
  time = "year",
  value = "value",
  distributions = c("gev", "gpa", "pe3", "ln2", "ln3", "gum"),
  methods = c("lmom"),
  return_period = c(10, 25, 50, 100),
  gof = TRUE,
  diagnostics = TRUE,
  select = "ai"
)

batch
```

# Station metadata

The `stations` table summarizes the samples processed by the batch workflow. Batch result tables print compactly in the console. Use `as.data.frame()` when you need the complete table for programming, export, or reporting.

```{r stations}
stations <- alea_results(batch, "stations")
stations
as.data.frame(stations)
```

# Fit summaries

The `fits` table records one row per attempted station-distribution-method fit. Successful fits are stored as full `alea_fit` objects in the `fit_objects` component.

```{r fits}
fits <- alea_results(batch, "fits")
head(fits)
```

```{r fit-objects}
fit_objects <- alea_results(batch, "fit_objects")
length(fit_objects)
```

# AI-assisted selected models

When `select = "ai"`, ALEA-R applies AI-assisted distribution-selection support at the station level. The selected distribution family is then resolved to an available fitted model using the method-priority rule.

```{r selected-models}
selected_models <- alea_results(batch, "selected_models")
selected_models
```

FADS_AI output should be interpreted as model-based decision-support evidence for candidate distribution families. It is not proof of the true generating distribution and should not replace goodness-of-fit assessment, diagnostics, quantile uncertainty evaluation, or hydrological judgement.

# Quantiles

When `return_period` is supplied, the batch object contains a flat quantile table for successful fitted models and requested return periods.

```{r quantiles}
quantiles <- alea_results(batch, "quantiles")
quantiles
head(as.data.frame(quantiles))
```

# Goodness-of-fit results

When `gof = TRUE`, the batch object contains goodness-of-fit statistics and information criteria for successful fitted models.

```{r gof}
gof <- alea_results(batch, "gof")
gof
head(as.data.frame(gof))
```

Goodness-of-fit results summarize distribution fit from different perspectives. They should be interpreted together with diagnostics, quantile uncertainty, graphical checks, and hydrological judgement.

# Diagnostics

When `diagnostics = TRUE`, the batch object contains sample-level diagnostics for the fitted station samples.

```{r diagnostics}
diagnostics <- alea_results(batch, "diagnostics")
diagnostics
head(as.data.frame(diagnostics))
```

Diagnostics are intended to flag possible data-quality issues or assumption concerns. Diagnostic warnings do not automatically invalidate a fitted model.

# Structured errors

Batch workflows use structured error capture. A failure for one station, distribution, method, quantile calculation, goodness-of-fit calculation, diagnostic calculation, or AI-selection step does not stop the full workflow.

Errors can be inspected with:

```{r errors}
errors <- alea_results(batch, "errors")
errors
```

When no errors are recorded, the errors table is still returned as an empty data frame with stable columns. This makes downstream export and reporting workflows easier to automate.

# Batch plots

The `plot()` method for `alea_batch` objects provides summaries for selected models, quantiles, goodness-of-fit results, and diagnostics.

```{r plot-selected-models}
plot(batch, type = "selected_models")
```

```{r plot-quantiles}
plot(batch, type = "quantiles")
```

```{r plot-gof}
plot(batch, type = "gof", statistic = "aic")
```

```{r plot-diagnostics-status}
plot(batch, type = "diagnostics")
```

For one hypothesis-test diagnostic, use the `diagnostic` argument.

```{r plot-single-diagnostic}
plot(batch, type = "diagnostics", diagnostic = "stationarity")
```

# Exporting batch results

The general export helper can export flat batch tables. Object-list fields such as `fit_objects` and `selection_objects` are intentionally not exported as CSV tables by default.

```{r export-batch, eval = FALSE}
alea_export(batch, path = "batch_results", type = "all")
```

# Recommended interpretation workflow

For each station, a practical workflow is:

1. Review the station metadata and data-quality diagnostics.
2. Check which candidate models fitted successfully.
3. Compare GOF statistics and information criteria.
4. Review AI-assisted distribution-selection support.
5. Inspect quantile estimates and plots.
6. Make the final modelling decision using statistical evidence and hydrological judgement.
