---
title: "Goodness-of-fit, diagnostics, and AI-assisted selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Goodness-of-fit, diagnostics, and AI-assisted selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This vignette shows how to combine goodness-of-fit statistics, sample
 diagnostics, and AI-assisted distribution-selection support in ALEA-R.

These tools answer different questions:

- goodness-of-fit statistics summarize how fitted distributions compare with
  the observed sample;
- diagnostics flag possible data-quality or assumption concerns;
- FADS_AI provides model-based support for candidate distribution families.

No single output should be interpreted in isolation. FADS_AI output is
 decision-support evidence, not proof of the true generating distribution.
Final model selection should also consider hydrological judgement, sample
quality, return-level uncertainty, and the purpose of the analysis.

# Example sample

The example below uses a small synthetic annual-maximum-like sample. In real
applications, use a quality-controlled hydrological series.

```{r setup}
library(ALEA)

x <- c(
  42.1, 38.5, 51.3, 47.0, 62.4,
  55.2, 49.8, 58.1, 60.3, 45.7,
  66.8, 53.9, 71.2, 64.0, 57.6,
  69.5, 73.1, 61.7, 52.4, 75.8,
  68.2, 59.9, 77.4, 81.0, 72.6
)
```

# Fit candidate distributions

For illustration, attempt to fit the six distributions supported in the initial
ALEA-R implementation using L-moments. Some distribution/sample combinations may
be invalid for a specific estimation method, so this vignette keeps only
successful fits for the comparison tables and plots.

```{r fit-candidates}
distributions <- c("gev", "gpa", "pe3", "ln2", "ln3", "gum")

fit_one <- function(dist) {
  tryCatch(
    alea_fit(x, distribution = dist, method = "lmom"),
    error = function(e) NULL
  )
}

fits <- lapply(distributions, fit_one)
names(fits) <- distributions
fits <- Filter(Negate(is.null), fits)

names(fits)
```

# Goodness-of-fit statistics

`alea_gof()` computes empirical distribution function statistics and
information criteria for one fitted model.

```{r gof-single}
reference_fit <- fits[[1]]
gof_reference <- alea_gof(reference_fit)
gof_reference
```

The default output includes:

- Kolmogorov-Smirnov statistic (`ks`);
- Cramer-von Mises statistic (`cvm`);
- Anderson-Darling statistic (`ad`);
- log-likelihood (`loglik`);
- Akaike information criterion (`aic`);
- Bayesian information criterion (`bic`).

Calibrated goodness-of-fit p-values and chi-square goodness-of-fit tests are
not part of the initial implementation.

```{r plot-gof-single}
plot(gof_reference, type = "statistic")
```

# Compare candidate distributions with GOF

A practical workflow is to compute goodness-of-fit statistics for all candidate
models and inspect the results together.

```{r gof-all}
gof_all <- do.call(rbind, lapply(fits, alea_gof))
gof_all
```

The rank plot helps compare candidate models across statistics. Rank 1
represents the best model for each statistic.

```{r plot-gof-rank}
plot(gof_all, type = "rank")
```

A distribution that ranks well across several statistics may deserve further
attention. However, goodness-of-fit statistics should not be treated as the only
basis for final model selection, especially for short hydrological records.

# Sample diagnostics

Diagnostics evaluate the sample and fitted-object data context. They are not
replacement goodness-of-fit tests.

```{r diagnostics}
diagnostics <- alea_diagnostics(reference_fit)
diagnostics
```

Simple diagnostics check sample size, missing or non-finite values, ties, range
adequacy, and skewness. When the suggested package `trend` is available,
hypothesis-based diagnostics can also evaluate randomness, independence,
homogeneity, and stationarity.

```{r plot-diagnostics-status}
plot(diagnostics, type = "status")
```

For hypothesis-test diagnostics, p-values can also be displayed directly.

```{r plot-diagnostics-pvalue}
plot(diagnostics, type = "p_value")
```

Diagnostic warnings should trigger review. They do not automatically invalidate
a fitted model. For example, a warning about sample size may indicate that
return-level uncertainty should be interpreted more cautiously.

# AI-assisted distribution selection

`alea_select()` applies the bundled FADS_AI lightweight operational application
model to the observed sample and returns model-based support for candidate
distribution families.

```{r ai-selection}
selection <- alea_select(x)
selection
```

The ranking table shows the model-based support assigned to each candidate
family.

```{r ai-ranking}
as.data.frame(selection)
```

```{r plot-ai-selection}
plot(selection)
```

FADS_AI probabilities should be reported as model-based support, not as proof
of the true distribution. The top-ranked family should be interpreted together
with diagnostics, GOF behavior, return-level uncertainty, and hydrological
knowledge.

# Combining evidence

A cautious workflow is:

1. Check diagnostics before interpreting model fits.
2. Compare several candidate distributions with GOF statistics.
3. Review FADS_AI support and the top1-top2 probability margin.
4. Estimate return levels and uncertainty for plausible candidates.
5. Use hydrological judgement before choosing a final model.

The following compact table joins three types of evidence: AIC ranking,
AI-assisted support, and diagnostic status for the sample used in the fitted
object.

```{r compact-evidence}
aic_table <- subset(gof_all, statistic == "aic")
aic_table <- aic_table[order(aic_table$estimate), c("distribution", "method", "estimate")]
names(aic_table)[names(aic_table) == "estimate"] <- "aic"
aic_table$aic_rank <- seq_len(nrow(aic_table))

ai_table <- as.data.frame(selection)[, c("distribution", "probability", "rank")]
names(ai_table) <- c("distribution", "ai_probability", "ai_rank")

merge(aic_table, ai_table, by = "distribution", all.x = TRUE)
```

This type of table is useful for reporting, but it should not be reduced to an
automatic rule. Differences among candidate distributions can be small, and
uncertainty may be large for high return periods.

# Return-level uncertainty for plausible candidates

After narrowing candidate distributions, inspect return levels and uncertainty.
The example below uses the reference fit selected above. In a full analysis,
repeat this step for other plausible candidates.

```{r return-level-ci}
ci <- confint(
  reference_fit,
  parm = "return_level",
  return_period = c(10, 25, 50, 100),
  level = 0.95,
  method = "bootstrap",
  n_boot = 50,
  seed = 123
)

ci
```

```{r plot-return-level-ci}
plot(ci)
```

The small `n_boot` value is used only to keep the vignette lightweight. For
substantive analyses, use a larger number of bootstrap resamples and assess the
stability of the interval estimates.

# Reporting guidance

When reporting results, prefer language such as:

> The FADS_AI model assigned the highest support to one candidate family for
> this sample. GOF statistics and diagnostics were reviewed together with
> return-level uncertainty before selecting the final model.

Avoid statements such as:

> The AI proved that the true distribution is GEV.

# Summary

Goodness-of-fit statistics, diagnostics, AI-assisted selection, and return-level
uncertainty are complementary. ALEA-R is designed to keep these outputs
structured and reportable, while leaving the final modelling decision to the
analyst.
