---
title: "Plots and exports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plots and exports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

ALEA-R plot methods return `ggplot` objects. This makes plots easy to print,
customize, combine with ordinary `ggplot2` workflows, and save to files.

The plotting layer covers fitted distributions, quantiles, bootstrap
confidence intervals, goodness-of-fit summaries, diagnostics, AI-assisted
selection, and batch results. The export layer provides helpers for saving one
plot, several plots, data frames, and flat batch result tables.

```{r setup}
library(ALEA)
```

# Example data

This vignette uses a small synthetic annual-maximum-like sample. The values are
used only to demonstrate package workflows.

```{r data}
x <- c(
  42.1, 38.5, 51.3, 47.0, 62.4,
  55.2, 49.8, 58.1, 60.3, 45.7,
  67.8, 52.6, 44.9, 57.5, 63.1,
  70.4, 59.0, 48.7, 54.3, 61.9,
  66.2, 50.5, 46.8, 64.7, 72.1
)

fit <- alea_fit(x, distribution = "gev", method = "lmom")
fit
```

# Fitted-model plots

`plot.alea_fit()` supports density, CDF, Q-Q, P-P, and quantile plots.

```{r fit-density}
p_density <- plot(fit, type = "density")
p_density
```

```{r fit-cdf}
p_cdf <- plot(fit, type = "cdf")
p_cdf
```

```{r fit-qq}
p_qq <- plot(fit, type = "qq")
p_qq
```

```{r fit-pp}
p_pp <- plot(fit, type = "pp")
p_pp
```

```{r fit-quantile}
p_fit_rl <- plot(
  fit,
  type = "quantile",
  return_period = c(2, 5, 10, 25, 50, 100)
)
p_fit_rl
```

# Quantile plots

Quantile estimates are represented by `alea_quantile` objects.

```{r quantile-object}
rl <- alea_quantile(
  fit,
  return_period = c(2, 5, 10, 25, 50, 100)
)

rl
```

```{r quantile-plot}
p_rl <- plot(rl)
p_rl
```

Quantile plots use a Gumbel return-period axis by default. The axis can also
be displayed on a logarithmic or linear scale.

```{r quantile-plot-log}
plot(rl, return_period_scale = "log")
```

# Bootstrap confidence-interval plots

Quantile confidence intervals are represented by `alea_quantile_ci`
objects. This example uses a small number of bootstrap resamples to keep the
vignette fast. Real analyses should use a larger value of `n_boot`.

```{r quantile-ci}
ci <- confint(
  fit,
  parm = "quantile",
  return_period = c(10, 25, 50, 100),
  level = 0.95,
  method = "bootstrap",
  n_boot = 50,
  seed = 123
)

ci
```

```{r quantile-ci-plot}
p_ci <- plot(ci)
p_ci
```

# Goodness-of-fit plots

`alea_gof()` reports empirical distribution function statistics and information
criteria for fitted models.

```{r gof-object}
gof <- alea_gof(fit)
gof
```

```{r gof-statistic-plot}
p_gof_stat <- plot(gof, type = "statistic")
p_gof_stat
```

To demonstrate the GOF rank plot, fit several candidate distributions and keep
only successful fits.

```{r gof-rank-data}
distributions <- c("gev", "gpa", "pe3", "ln2", "ln3", "gum")

fit_one <- function(dist) {
  tryCatch(
    alea_fit(x, distribution = dist, method = "lmom"),
    error = function(e) NULL
  )
}

fits <- lapply(distributions, fit_one)
names(fits) <- distributions
fits <- Filter(Negate(is.null), fits)

gof_all <- do.call(
  rbind,
  lapply(fits, function(obj) as.data.frame(alea_gof(obj)))
)
class(gof_all) <- c("alea_gof", "data.frame")

unique(gof_all$distribution)
```

```{r gof-rank-plot}
p_gof_rank <- plot(gof_all, type = "rank")
p_gof_rank
```

# Diagnostic plots

Diagnostics are sample-level checks. They help identify possible data-quality
issues or assumption concerns, but diagnostic warnings do not automatically
invalidate a fitted model.

```{r diagnostics-object}
diagnostics <- alea_diagnostics(fit)
diagnostics
```

```{r diagnostics-status-plot}
p_diag_status <- plot(diagnostics, type = "status")
p_diag_status
```

```{r diagnostics-p-value-plot}
p_diag_p <- plot(diagnostics, type = "p_value")
p_diag_p
```

# AI-assisted selection plot

`alea_select()` returns model-based support for candidate distribution families.
FADS_AI output is decision-support evidence, not proof of the true generating
distribution.

```{r selection-object}
selection <- alea_select(x)
selection
```

```{r selection-plot}
p_selection <- plot(selection)
p_selection
```

# Batch plots

Batch analysis produces an `alea_batch` object with flat result tables and full
S3 objects for successful station-level workflows.

```{r batch-data}
batch_data <- data.frame(
  station = rep(c("A", "B"), each = length(x)),
  year = rep(seq_along(x), times = 2),
  value = c(
    x,
    x * 1.20 + c(
      0.3, -0.2, 0.1, 0.4, -0.1,
      0.2, -0.3, 0.1, 0.5, -0.2,
      0.0, 0.3, -0.4, 0.2, 0.1,
      -0.1, 0.4, -0.2, 0.2, 0.3,
      -0.3, 0.1, 0.2, -0.1, 0.4
    )
  )
)

batch <- alea_batch_fit(
  data = batch_data,
  station = "station",
  time = "year",
  value = "value",
  distributions = c("gev", "gum", "ln2"),
  methods = "lmom",
  return_period = c(10, 25, 50, 100),
  gof = TRUE,
  diagnostics = TRUE,
  select = "ai"
)

batch
```

```{r batch-selected-models-plot}
p_batch_selected <- plot(batch, type = "selected_models")
p_batch_selected
```

```{r batch-quantiles-plot}
p_batch_rl <- plot(batch, type = "quantiles")
p_batch_rl
```

```{r batch-gof-plot}
p_batch_gof <- plot(batch, type = "gof", statistic = "aic")
p_batch_gof
```

```{r batch-diagnostics-plot}
p_batch_diag <- plot(batch, type = "diagnostics")
p_batch_diag
```

# Saving one plot

Use `alea_save_plot()` to save one `ggplot` object. In this vignette, files are
written to a temporary directory so the example does not leave files behind.

```{r save-one-plot}
out_dir <- tempdir()

one_plot_file <- file.path(out_dir, "alea_quantile.png")

if (file.exists(one_plot_file)) {
  unlink(one_plot_file)
}

alea_save_plot(
  p_rl,
  filename = one_plot_file,
  width = 7,
  height = 5,
  dpi = 300
)

file.exists(one_plot_file)
```

# Saving several plots

Use `alea_save_plots()` to save a named list of plots.

```{r save-several-plots}
plot_files <- alea_save_plots(
  plots = list(
    density = p_density,
    quantile = p_rl,
    selection = p_selection
  ),
  directory = out_dir,
  prefix = "alea_example",
  extension = "png",
  width = 7,
  height = 5,
  dpi = 300,
  overwrite = TRUE
)

plot_files
file.exists(plot_files)
```

# Exporting data frames

Use `alea_export()` to export data frames as CSV files.

```{r export-data-frame}
rl_file <- file.path(out_dir, "quantiles.csv")

alea_export(
  as.data.frame(rl),
  path = rl_file,
  overwrite = TRUE
)

file.exists(rl_file)
```

# Exporting batch result tables

`alea_export()` can also export non-empty flat tables from an `alea_batch`
object. Heavy object-list fields such as `fit_objects` and `selection_objects`
are intentionally not exported as CSV tables.

```{r export-batch}
batch_dir <- file.path(out_dir, "alea_batch_export")

if (!dir.exists(batch_dir)) {
  dir.create(batch_dir, recursive = TRUE)
}

action <- alea_export(
  batch,
  path = batch_dir,
  type = "all",
  overwrite = TRUE
)

action
list.files(batch_dir)
```

# Summary

The ALEA-R plotting and export layer is designed for reporting workflows:

- plot methods return `ggplot` objects;
- quantile plots use a Gumbel return-period axis by default;
- bootstrap confidence intervals can be plotted directly;
- GOF, diagnostics, AI selection, and batch summaries have dedicated plots;
- export helpers support plots, data frames, and flat batch result tables.
