---
title: "Single-site frequency analysis with ALEA-R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-site frequency analysis with ALEA-R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This vignette shows a complete single-site hydrological frequency-analysis
workflow with ALEA-R.

The workflow includes:

- preparing an observed sample;
- fitting a probability distribution;
- estimating return levels;
- computing bootstrap confidence intervals;
- assessing goodness-of-fit;
- checking sample diagnostics;
- obtaining AI-assisted distribution-selection support;
- creating plots for reporting.

ALEA-R uses S3 objects for user-facing results. The examples below use the
initial candidate distributions supported by the package: `gev`, `gpa`, `pe3`,
`ln2`, `ln3`, and `gum`.

# Setup

```{r setup}
library(ALEA)
```

# Example sample

For this vignette, we use a small synthetic annual-maximum-like sample. In a real
application, `x` should be replaced by an observed hydrological series such as
annual maximum rainfall, annual peak discharge, or another frequency-analysis
variable.

```{r sample-data}
x <- c(
  42.1, 38.5, 51.3, 47.0, 62.4,
  55.2, 49.8, 58.1, 60.3, 45.7,
  66.8, 53.4, 57.9, 63.2, 70.5,
  48.6, 52.7, 59.4, 61.8, 56.1,
  73.2, 67.4, 64.9, 69.1, 75.6
)

summary(x)
```

# Fit one distribution

The main fitting function is `alea_fit()`. It fits one supported probability
distribution to one numeric sample using one estimation method.

Here we fit a GEV distribution with L-moment estimation.

```{r fit-gev}
fit <- alea_fit(
  x,
  distribution = "gev",
  method = "lmom"
)

fit
coef(fit)
```

The fitted object stores the original sample and fitted parameters. Downstream
workflows use this object directly, avoiding repeated specification of the
sample, distribution, and estimation method.

# Return levels

Return levels are estimated with `alea_return_level()`. The argument
`return_period` defines the return periods of interest.

```{r return-levels}
return_period <- c(2, 5, 10, 25, 50, 100)

rl <- alea_return_level(
  fit,
  return_period = return_period
)

rl
```

The returned object is an `alea_return_level` data frame. It can be converted to
a plain data frame when needed.

```{r return-levels-data-frame}
as.data.frame(rl)
```

# Bootstrap confidence intervals

Return-level confidence intervals are computed with the S3 `confint()` method
for `alea_fit` objects.

The initial implementation provides percentile bootstrap confidence intervals
for return levels. Parameter confidence intervals, asymptotic return-level
intervals, and generic delta-method intervals are not implemented in the initial
release.

The example uses a small number of bootstrap resamples to keep vignette checks
fast. For applied analyses, use a larger value such as `n_boot = 500` or more,
depending on the required precision and available computing time.

```{r bootstrap-ci}
ci <- confint(
  fit,
  parm = "return_level",
  return_period = return_period,
  level = 0.95,
  method = "bootstrap",
  n_boot = 50,
  seed = 123
)

ci
```

The output records the requested number of bootstrap resamples and the number of
successful and failed resamples.

```{r bootstrap-ci-data-frame}
as.data.frame(ci)
```

# Goodness-of-fit

Goodness-of-fit statistics are computed with `alea_gof()`.

```{r gof}
gof <- alea_gof(fit)

gof
```

The initial goodness-of-fit workflow reports empirical distribution function
statistics and information criteria, including Kolmogorov-Smirnov,
Cramer-von Mises, Anderson-Darling, log-likelihood, AIC, and BIC.

Calibrated goodness-of-fit p-values and chi-square goodness-of-fit tests are
deferred in the initial implementation.

```{r gof-data-frame}
as.data.frame(gof)
```

# Sample diagnostics

Sample diagnostics are computed with `alea_diagnostics()`. Diagnostics are
sample-level checks intended to flag possible data-quality issues or assumption
concerns. They are not distribution-specific goodness-of-fit tests.

```{r diagnostics}
diagnostics <- alea_diagnostics(fit)

diagnostics
```

Diagnostic warnings do not automatically invalidate a fitted model. They should
be reviewed together with goodness-of-fit results, return-level uncertainty, and
hydrological judgement.

```{r diagnostics-data-frame}
as.data.frame(diagnostics)
```

# AI-assisted distribution selection

ALEA-R includes AI-assisted distribution-selection support through `alea_select()`.
The default engine is the bundled FADS_AI lightweight operational application
model.

```{r ai-selection}
selection <- alea_select(x)

selection
```

The ranking table gives model-based support for each candidate distribution.

```{r ai-ranking}
as.data.frame(selection)
```

FADS_AI output should be interpreted as decision-support evidence for candidate
distribution families. It is not proof of the true generating distribution and
should not replace goodness-of-fit assessment, diagnostics, return-level
uncertainty evaluation, or hydrological judgement.

# Plots

ALEA-R plot methods return `ggplot` objects, so they can be displayed,
customized, and saved with standard ggplot2 workflows.

## Fitted distribution plots

```{r plot-density}
plot(fit, type = "density")
```

```{r plot-cdf}
plot(fit, type = "cdf")
```

```{r plot-qq}
plot(fit, type = "qq")
```

```{r plot-pp}
plot(fit, type = "pp")
```

## Return-level plots

```{r plot-return-level-fit}
plot(fit, type = "return_level", return_period = return_period)
```

```{r plot-return-level-object}
plot(rl)
```

```{r plot-return-level-ci}
plot(ci)
```

Return-level plots use a Gumbel return-period axis by default. The scale can be
changed with `return_period_scale = "log"` or `return_period_scale = "linear"`.

```{r plot-return-level-log-scale}
plot(rl, return_period_scale = "log")
```

## Goodness-of-fit and diagnostics plots

```{r plot-gof-statistic}
plot(gof, type = "statistic")
```

```{r plot-diagnostics-status}
plot(diagnostics, type = "status")
```

```{r plot-diagnostics-p-value}
plot(diagnostics, type = "p_value")
```

## AI-selection plot

```{r plot-ai-selection}
plot(selection)
```

# Exporting results

Plot objects can be saved with `alea_save_plot()`.

```{r save-plot, eval = FALSE}
p <- plot(fit, type = "return_level", return_period = return_period)

alea_save_plot(
  p,
  filename = "single_site_return_level.png",
  width = 7,
  height = 5,
  dpi = 300
)
```

Data-frame-like outputs can be exported with `alea_export()`.

```{r export-tables, eval = FALSE}
alea_export(rl, path = "single_site_return_levels.csv")
alea_export(as.data.frame(gof), path = "single_site_gof.csv")
alea_export(as.data.frame(diagnostics), path = "single_site_diagnostics.csv")
```

# Interpretation

A single-site frequency analysis should not rely on one diagnostic or one model
selection criterion alone. In practice, the fitted distribution should be
reviewed together with:

- the fitted density, CDF, Q-Q, P-P, and return-level plots;
- return-level uncertainty;
- goodness-of-fit statistics and information criteria;
- sample diagnostics;
- AI-assisted distribution-selection support;
- hydrological knowledge of the site and process.

The FADS_AI ranking provides model-based decision-support evidence for the
candidate families. It is not proof of the true generating distribution.
